---
title: "Test R Markdown Document"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

RENDER_MAP          <- TRUE
RENDER_COVERAGE_MAP <- TRUE
```

## Extracting Surveillance Cameras from OpenStreetMap

This document demonstrates how to query OpenStreetMap for surveillance cameras in Chicago using `osmdata`.

```{r load-libraries}
library(osmdata)
library(sf)
library(terra)
library(stars)
library(gstat)
library(tidyverse)
library(leaflet)
library(tigris)
library(tidycensus)
library(DT)
library(crosstalk)
```

```{r get-chicago-bbox}
# Get the Chicago bounding box from OSM
chicago_bb <- getbb("Chicago, Illinois, USA")

# Build a Chicago bounding box polygon
chicago_bbox <- st_as_sfc(st_bbox(c(xmin = chicago_bb[1,1], ymin = chicago_bb[2,1], xmax = chicago_bb[1,2], ymax = chicago_bb[2,2]), crs = st_crs(4326)))
```

```{r get-tract-data}
readRenviron(".Renviron")
census_api_key(Sys.getenv("CENSUS_API_KEY"))

# Get census tract boundaries for Cook County (Chicago area)
cook_tracts <- tracts(state = "IL", county = "031", year = 2020)

# Get median household income data for tracts in Cook County
income_data <- get_acs(
  geography = "tract",
  variables = "B19013_001",  # Median household income
  state = "IL",
  county = "031",
  year = 2022,
  geometry = FALSE
)

# Join income data to tract boundaries
tract_income <- cook_tracts %>%
  left_join(income_data, by = c("GEOID" = "GEOID")) %>%
  filter(!is.na(estimate)) %>%
  st_transform(4326)

# Filter to Chicago area (approximate bounding box)
chicago_tracts <- st_filter(tract_income, chicago_bbox)

print(paste("Tracts in Chicago area:", nrow(chicago_tracts)))
```

```{r get-chicago-cameras}
cameras <- st_read("data/cameras.geojson.gz", quiet = TRUE)
cameras <- st_transform(cameras, crs = 4326)
cameras_chicago <- st_filter(cameras, chicago_bbox)
print(paste("Total ALPR cameras in Chicago bbox:", nrow(cameras_chicago)))
```

```{r get-red-light-cameras}
redlights_raw <- read.csv("data/redlights.csv")
redlights <- st_as_sf(
  redlights_raw, coords = c("LONGITUDE", "LATITUDE"), crs = 4326
)
print(paste("Total red light cameras:", nrow(redlights)))
```

```{r get-speed-cameras}
speed_cameras_raw <- read.csv("data/speed_cameras.csv")
speed_cameras <- st_as_sf(
  speed_cameras_raw, coords = c("LONGITUDE", "LATITUDE"), crs = 4326
)
print(paste("Total speed cameras:", nrow(speed_cameras)))
```

```{r get-pod-cameras}
pod_cameras <- st_read("data/pod_cameras.geojson", quiet = TRUE)
pod_cameras <- st_transform(pod_cameras, crs = 4326)
print(paste("Total POD cameras:", nrow(pod_cameras)))
```
Credit to [redshiftzero's Police Surveillance project](https://github.com/lucyparsons/policesurveillance/blob/gh-pages/pod.html) for POD camera locations

```{r pod-buffers}
# 0.25 mi = 1320 ft = 402.336 m; project to UTM 16N for metric buffering, then back to 4326
pod_buffers <- pod_cameras |>
  st_transform(32616) |>
  st_buffer(402.336) |>
  st_transform(4326)
```

```{r join-camera-tracts}
tract_cols <- chicago_tracts[, c("TRACTCE", "GEOID")]

cameras_chicago <- st_join(cameras_chicago, tract_cols)
redlights       <- st_join(redlights,       tract_cols)
speed_cameras   <- st_join(speed_cameras,   tract_cols)
pod_cameras     <- st_join(pod_cameras,     tract_cols)
```

```{r shared-camera-data}
coords_xy <- function(sf_obj) {
  xy <- st_coordinates(st_transform(sf_obj, 4326))
  data.frame(Lat = round(xy[, 2], 5), Lon = round(xy[, 1], 5))
}

camera_df <- bind_rows(
  cameras_chicago |>
    bind_cols(coords_xy(cameras_chicago)) |>
    st_drop_geometry() |>
    transmute(
      Type     = "ALPR",
      Location = coalesce(operator, "Unknown"),
      Details  = coalesce(brand, ""),
      Tract    = coalesce(TRACTCE, "—"),
      Lat, Lon,
      popup = paste0(
        "<b>ALPR Camera</b><br>",
        "Operator: ", coalesce(operator, "Unknown"),
        "<br>Brand: ", coalesce(brand, "—"),
        "<br>Zone: ", coalesce(surveillanceZone, "—"),
        "<br>Census Tract: ", coalesce(TRACTCE, "unknown")
      )
    ),
  redlights |>
    bind_cols(coords_xy(redlights)) |>
    st_drop_geometry() |>
    transmute(
      Type     = "Red Light",
      Location = INTERSECTION,
      Details  = paste0("Go Live: ", GO.LIVE.DATE),
      Tract    = coalesce(TRACTCE, "—"),
      Lat, Lon,
      popup = paste0(
        "<b>Red Light Camera</b><br>",
        "Intersection: ", INTERSECTION, "<br>",
        "Approaches: ", FIRST.APPROACH,
        ifelse(!is.na(SECOND.APPROACH) & SECOND.APPROACH != "",
               paste0(", ", SECOND.APPROACH), ""),
        ifelse(!is.na(THIRD.APPROACH) & THIRD.APPROACH != "",
               paste0(", ", THIRD.APPROACH), ""), "<br>",
        "Go Live: ", GO.LIVE.DATE, "<br>",
        "Census Tract: ", coalesce(TRACTCE, "unknown")
      )
    ),
  speed_cameras |>
    bind_cols(coords_xy(speed_cameras)) |>
    st_drop_geometry() |>
    transmute(
      Type     = "Speed",
      Location = ADDRESS,
      Details  = paste0("Go Live: ", GO.LIVE.DATE),
      Tract    = coalesce(TRACTCE, "—"),
      Lat, Lon,
      popup = paste0(
        "<b>Speed Camera</b><br>",
        "Address: ", ADDRESS, "<br>",
        "Approaches: ", FIRST.APPROACH,
        ifelse(!is.na(SECOND.APPROACH) & SECOND.APPROACH != "",
               paste0(", ", SECOND.APPROACH), ""), "<br>",
        "Go Live: ", GO.LIVE.DATE, "<br>",
        "Census Tract: ", coalesce(TRACTCE, "unknown")
      )
    ),
  pod_cameras |>
    bind_cols(coords_xy(pod_cameras)) |>
    st_drop_geometry() |>
    transmute(
      Type     = "POD",
      Location = street,
      Details  = paste0("Camera #", camera_number),
      Tract    = coalesce(TRACTCE, "—"),
      Lat, Lon,
      popup = paste0(
        "<b>POD Camera</b><br>",
        "Camera #: ", camera_number, "<br>",
        "Street: ", street, "<br>",
        "Zip: ", zip, "<br>",
        "Census Tract: ", coalesce(TRACTCE, "unknown")
      )
    )
)

camera_df$Type <- factor(camera_df$Type, levels = c("ALPR", "Red Light", "Speed", "POD"))

shared_cameras <- SharedData$new(camera_df)
```

```{r render-map}
income_pal <- colorQuantile("YlOrRd", chicago_tracts$estimate, n = 5, na.color = "#cccccc")

tract_popup <- paste0(
  "<b>Census Tract </b>", chicago_tracts$TRACTCE, "<br>",
  "Median Household Income: $", formatC(chicago_tracts$estimate, format = "f", digits = 0, big.mark = ","), "<br>",
  "Margin of Error: ±$", formatC(chicago_tracts$moe, format = "f", digits = 0, big.mark = ",")
)

type_pal <- colorFactor(
  palette = c("blue", "orange", "green", "purple"),
  levels  = c("ALPR", "Red Light", "Speed", "POD")
)

m <- leaflet() |>
  addTiles() |>
  addPolygons(
    data = chicago_tracts,
    fillColor = ~income_pal(estimate),
    fillOpacity = 0.6,
    color = "white", weight = 0.5,
    popup = tract_popup,
    group = "Median Income"
  ) |>
  addPolygons(
    data = pod_buffers,
    fillColor = "purple", fillOpacity = 0.15,
    color = "purple", weight = 1, opacity = 0.4,
    group = "POD Buffers"
  ) |>
  addCircleMarkers(
    data = camera_df[camera_df$Type == "ALPR", ],
    lat = ~Lat, lng = ~Lon,
    radius = 4, color = type_pal("ALPR"), stroke = FALSE, fillOpacity = 0.8,
    popup = ~popup,
    group = "ALPR"
  ) |>
  addCircleMarkers(
    data = camera_df[camera_df$Type == "Red Light", ],
    lat = ~Lat, lng = ~Lon,
    radius = 4, color = type_pal("Red Light"), stroke = FALSE, fillOpacity = 0.8,
    popup = ~popup,
    group = "Red Light"
  ) |>
  addCircleMarkers(
    data = camera_df[camera_df$Type == "Speed", ],
    lat = ~Lat, lng = ~Lon,
    radius = 4, color = type_pal("Speed"), stroke = FALSE, fillOpacity = 0.8,
    popup = ~popup,
    group = "Speed"
  ) |>
  addCircleMarkers(
    data = camera_df[camera_df$Type == "POD", ],
    lat = ~Lat, lng = ~Lon,
    radius = 4, color = type_pal("POD"), stroke = FALSE, fillOpacity = 0.8,
    popup = ~popup,
    group = "POD"
  ) |>
  addLayersControl(
    overlayGroups = c("Median Income", "POD Buffers", "ALPR", "Red Light", "Speed", "POD"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  addLegend(
    position = "bottomright",
    pal = income_pal,
    values = chicago_tracts$estimate,
    title = "Median Household<br>Income (Quintile)",
    labFormat = function(type, cuts, p) {
      n <- length(cuts)
      paste0(round(p[-n] * 100), "% – ", round(p[-1] * 100), "%")
    }
  ) |>
  addLegend(
    position = "bottomleft",
    pal = type_pal,
    values = c("ALPR", "Red Light", "Speed", "POD"),
    title = "Camera Type"
  ) |>
  setView(lng = mean(c(chicago_bb[1,])), lat = mean(c(chicago_bb[2,])), zoom = 11)

type_filter <- filter_checkbox(
  "type_filter", "Camera Type", shared_cameras, ~Type, inline = TRUE
)

dt <- datatable(
  shared_cameras,
  filter   = "top",
  rownames = FALSE,
  colnames = c("Type", "Operator", "Details", "Census Tract", "Lat", "Lon", ""),
  options  = list(
    pageLength = 25,
    order      = list(list(0, "asc")),
    columnDefs = list(
      list(visible = FALSE, targets = c(3, 4, 5, 6))
    )
  )
)

if (RENDER_MAP) {
  htmltools::browsable(htmltools::tagList(type_filter, m, dt))
}
```

## Data Summary

You can now analyze the three surveillance camera datasets by location, type, and other attributes.

### POD Camera Buffer Crime Analysis

```{r load-crimes}
crimes_raw <- read.csv("data/chicago_crimes.csv")

crimes <- crimes_raw |>
  mutate(Primary.Type = if_else(Primary.Type == "CRIM SEXUAL ASSAULT",
                                "CRIMINAL SEXUAL ASSAULT",
                                Primary.Type)) |>
  filter(!is.na(Latitude) & !is.na(Longitude)) |>
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

print(paste("Total crimes with valid coordinates:", nrow(crimes)))
```

```{r pod-buffer-crime-analysis}
# Union all POD buffers into one geometry for efficient intersection test
pod_buffer_union <- st_union(pod_buffers)

in_buffer <- st_within(crimes, pod_buffer_union, sparse = FALSE)[, 1]

n_total     <- nrow(crimes)
n_in_buffer <- sum(in_buffer)
pct         <- n_in_buffer / n_total * 100

cat(sprintf(
  "%d of %d crimes (%.1f%%) occurred within a quarter-mile of a POD camera.\n",
  n_in_buffer, n_total, pct
))

crimes |>
  st_drop_geometry() |>
  as_tibble() |>
  filter(in_buffer) |>
  count(Primary.Type, name = "n") |>
  mutate(pct = n / sum(n) * 100) |>
  arrange(desc(n)) |>
  print(n = Inf)
```

## Spatial Autocorrelation: Cameras vs. Crime (Tract Level)


```{r tract-aggregates}
# Pool all four camera types and count per tract
cameras_per_tract <- bind_rows(
  cameras_chicago |> st_drop_geometry() |> select(TRACTCE),
  redlights       |> st_drop_geometry() |> select(TRACTCE),
  speed_cameras   |> st_drop_geometry() |> select(TRACTCE),
  pod_cameras     |> st_drop_geometry() |> select(TRACTCE)
) |>
  filter(!is.na(TRACTCE)) |>
  count(TRACTCE, name = "n_cameras")

# Join crimes to tracts, then count
crimes_per_tract <- crimes |>
  st_join(tract_cols) |>
  st_drop_geometry() |>
  filter(!is.na(TRACTCE)) |>
  count(TRACTCE, name = "n_crimes")

# Attach counts to tract polygons; convert to density per km² to control for tract size
tract_metrics <- chicago_tracts |>
  left_join(cameras_per_tract, by = "TRACTCE") |>
  left_join(crimes_per_tract,  by = "TRACTCE") |>
  mutate(
    n_cameras     = replace_na(n_cameras, 0),
    n_crimes      = replace_na(n_crimes,  0),
    area_mi2      = as.numeric(st_area(geometry)) / 2589988.11,
    cam_density   = n_cameras / area_mi2,
    crime_density = n_crimes  / area_mi2
  )
```

```{r spatial-weights}
library(spdep)

# Queen contiguity neighbors with row-standardized weights;
# zero.policy keeps tracts with no neighbors (e.g., lakefront) in the analysis
nb <- poly2nb(tract_metrics, queen = TRUE)
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
```

```{r univariate-moran}
# Are cameras and crimes each spatially clustered?
moran_cam   <- moran.test(tract_metrics$cam_density,   lw, zero.policy = TRUE)
moran_crime <- moran.test(tract_metrics$crime_density, lw, zero.policy = TRUE)

print(moran_cam)
print(moran_crime)
```

```{r bivariate-moran}
# Bivariate Moran's I: does camera density at tract i correlate with
# crime density in its neighbors? Standardize, then correlate x with W·y.
x     <- scale(tract_metrics$cam_density)[, 1]
y     <- scale(tract_metrics$crime_density)[, 1]
lag_y <- lag.listw(lw, y, zero.policy = TRUE)

bv_I <- sum(x * lag_y, na.rm = TRUE) / sum(x^2, na.rm = TRUE)
cat(sprintf("Bivariate Moran's I (cameras ~ lagged crime): %.3f\n", bv_I))

# Local bivariate Moran for tract-level hot/cold-spot mapping
local_bv <- localmoran_bv(x, y, lw, nsim = 199, zero.policy = TRUE)
head(local_bv)
```

## Camera Coverage Map

```{r combine-cameras}
# Project to Illinois StatePlane (EPSG:3435, feet) so buffer/grid distances
# are in native foot units — mirrors the projection choice in Lab 7
all_cameras <- bind_rows(
  cameras_chicago |> transmute(type = "ALPR"),
  redlights       |> transmute(type = "Red Light"),
  speed_cameras   |> transmute(type = "Speed"),
  pod_cameras     |> transmute(type = "POD")
) |>
  st_transform(3435)
```

```{r chicago-boundary}
# Download the true Chicago city boundary and reproject to match cameras
illinois_places <- suppressMessages(suppressWarnings(tigris::places(state = "IL", cb = TRUE)))
chicago_sf <- illinois_places |>
  filter(NAME == "Chicago") |>
  st_transform(3435)
```

```{r coverage-grid}
# Create a 1000 ft grid over Chicago — mirrors Lab 7's grid setup exactly
chicago_blank_raster <- st_bbox(chicago_sf) |>
  st_as_stars(dx = 1000, dy = 1000) |>
  st_crop(chicago_sf)

# Assign placeholder zeros so the grid has valid values (mirrors Lab 7)
chicago_blank_raster$values <- 0

# Plot to confirm grid coverage and alignment with Chicago boundary
plot(chicago_blank_raster, reset = FALSE)
plot(st_geometry(chicago_sf), add = TRUE, border = "red", lwd = 2)

# Convert to terra for rasterization and focal operations
chicago_template <- rast(chicago_blank_raster)
chicago_vect     <- vect(chicago_sf)
```

```{r camera-count-raster}
# Rasterize camera locations as counts per 1000-ft cell —
# each cell receives the number of cameras that fall within it (0 if none)
camera_raster <- rasterize(vect(all_cameras), chicago_template, fun = "count", background = 0)

plot(camera_raster, reset = FALSE)
plot(st_geometry(chicago_sf), add = TRUE, border = "red", lwd = 2)
```

```{r camera-coverage-raster}
# Mask count raster to Chicago boundary and export
coverage_masked <- mask(camera_raster, chicago_vect)
names(coverage_masked) <- "camera_coverage"

writeRaster(coverage_masked, "output/camera_coverage_idw_masked.tif", overwrite = TRUE)
```

```{r camera-coverage-grid}
# Square grid over Chicago (~1000 cells), filtered to city boundary
coverage_grid <- st_make_grid(chicago_sf, n = c(32, 32), square = TRUE) |>
  st_sf() |>
  st_filter(chicago_sf)

# Count cameras per grid cell via spatial intersection (both in EPSG:3435)
coverage_grid$n_cameras <- lengths(st_intersects(coverage_grid, all_cameras))

coverage_grid <- st_transform(coverage_grid, 4326)

st_write(coverage_grid, "output/camera_coverage_grid.geojson", delete_dsn = TRUE)
```

```{r render-coverage-map}
if (RENDER_COVERAGE_MAP) {
  valid_vals <- coverage_grid$n_cameras

  if (all(valid_vals == 0)) {
    stop("All grid cells have zero cameras — check camera-count-raster chunk output.")
  }

  coverage_pal <- colorNumeric("YlOrRd", domain = range(valid_vals), na.color = "transparent")

  leaflet(coverage_grid) |>
    addTiles() |>
    addPolygons(
      fillColor   = ~coverage_pal(n_cameras),
      fillOpacity = 0.7,
      color = NA, weight = 0,
      popup = ~paste0("Cameras in cell: ", n_cameras)
    ) |>
    addLegend(
      position = "bottomright",
      pal      = coverage_pal,
      values   = valid_vals,
      title    = "Camera Count<br>per Grid Cell"
    )
}
```

